---
title: "Proj 2 P01 Data Analyst Script"
author: "Ashley Morrison"
output:
  html_document: default
  pdf_document: default
date: "2023-02-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/projectnb/bf528/users/group_7/project_2/code/analyst")

library(dplyr)
library(ggplot2)
library(ggbreak) 
#install.packages("kableExtra")
library(kableExtra)
```

__Part 1__

Load the file cuffdiff_out/gene_exp.diff into R. Sort the data frame so that the smallest q_values are at the top (hint: look at the order function). Produce a table of the top ten differentially expressed genes, with their names, FPKM values, log fold change, p-value, and q-value in your report.

Info on the cuffdiff package: http://cole-trapnell-lab.github.io/cufflinks/cuffdiff/
```{r}
gene_exprs <- read.delim("../../samples/analyst_data/gene_exp.diff")
head(gene_exprs)

## sorting data by q-value
sorted_gene_exprs <- gene_exprs %>% arrange(q_value)
head(sorted_gene_exprs)

## making table
table_data <- sorted_gene_exprs %>% select(gene,value_1,value_2, log2.fold_change., p_value,q_value)
colnames(table_data) <- c("Gene Name","P0 FPKM","Adult FPKM","L2FC", "p-value", "q-value")
head(table_data)

#seeing how many samples have the same smallest q_value:
sum(table_data$`q-value`== 0.000318974) #2319
nrow(table_data) #36329 genes total

#lots of genes have same smallest q-value, so sort by q-value and then by L2FC
table_data2 <- table_data[order(table_data[,"q-value"],-abs(table_data[,"L2FC"])),]
head(table_data2)

#remove -Inf samples from top ten table:
final_table_data <- table_data2 %>% filter(L2FC!=-Inf & L2FC!=Inf)
head(final_table_data)

write.table(final_table_data[1:10,], file="top_10_diff_expressed_genes_P0_adult.csv")
```
Formating Table for report:
```{r}
#sorted by L2FC:
final_table_data[1:10,] %>%
  kbl(caption = "") %>%
  kable_classic(full_width = F, html_font = "Cambria")

#not sorted by L2FC:
table_data[1:10,] %>%
  kbl(caption = "") %>%
  kable_classic(full_width = F, html_font = "Cambria")

```


__Part 2__

Produce a histogram of the log2.foldchange column for all genes with the hist function. Try specifying different values for the breaks argument to control the number of bars in the plot and pick one you like best.
```{r}
pdf(file="P0_v_Adult_L2FC.pdf",height = 3,width=8)

ggplot(sorted_gene_exprs,aes(log2.fold_change.))+
  geom_histogram(binwidth = 0.1,bins = 1000)+ #adjust binwidth for better visualization
  ylim(0,15000)+ #set y-axis limits
  scale_y_break(c(1000, 14500),ticklabels=c(14500,15000))+ #adjust y-axis breaks for better visualization
  ggtitle("Distribtution of L2FC Between P0 and Adult Samples")+
  xlab("L2FC")

dev.off()

#look for most common L2FC value:
sort(table(sorted_gene_exprs$log2.fold_change.), decreasing = TRUE)[1:5]
  
```


__Part 3__

Create a new data frame that contains only the genes where the last column, named significant, is equal to yes. The subset function is useful for this kind of task, for example if df is a data frame with a column named A, then we can write df.sub <- subset(df,A==”x”). Note the nominal p-value and q-value that these genes have in your report.
```{r}
#subset by only significant genes
signf_genes <- gene_exprs %>% filter(significant=="yes")
head(signf_genes)
nrow(signf_genes) #5193 genes

```

__Part 4__

Create a second histogram of the log2 fold change values only for significant genes. What do you notice?
```{r}
pdf(file="Signif_P0_v_Adult_L2FC.pdf",height = 3,width=8)
ggplot(signf_genes,aes(log2.fold_change.))+
  geom_histogram(binwidth = 0.1,bins = 1000)+
  ggtitle("Distribtution of significant L2FC between P0 and Adult Samples")+
  xlab("L2FC")
dev.off()

#find the smallest still significant LF2C values and corresponding genes
sort((signf_genes$log2.fold_change.[signf_genes$log2.fold_change.>0]))[1:4]
sort((signf_genes$log2.fold_change.[signf_genes$log2.fold_change.<0]), decreasing = TRUE)[1:4]

signf_genes[signf_genes$log2.fold_change.==0.486609,] #Fh1
signf_genes[signf_genes$log2.fold_change.==-0.490812,] #Rpl28



```

__Part 5__

Further subset the significant gene data frame you just created into two separate data frames with only the up- and down-regulated genes using the log2.foldchange column. Include the number of up and down regulated genes in your report.

```{r}
#filter by genes with positive values (up-regulated in adult compared to P0)
signif_up <- signf_genes %>% filter(log2.fold_change. > 0) %>% arrange(q_value)
nrow(signif_up)

#filter by genes with negative values (down-regulated in adult compared to P0)
signif_down <- signf_genes %>% filter(log2.fold_change. < 0) %>% arrange(q_value)
nrow(signif_down)
```

__Part 6__

Using the write function, write out the up- and down- regulated gene names to separate files. The files should have one gene name on each line (leave genes that are separated by commas on the same line as is). Copy these files to your laptop or desktop, as we will be uploading them to a web application to perform gene set analysis.
```{r}
write.csv(signif_up[,"gene"],"up_regulated_genes.csv") #pull just gene column for up-regulated genes
write.csv(signif_down[,"gene"],"down_regulated_genes.csv") #pull just gene column for down-regulated genes
```

Making table of top GO terms:
```{r}
top_GO <- read.csv("top3_clusters.csv")

colnames(top_GO)[colnames(top_GO)=='X..of.total.Genes'] <- "Percent of Total Genes"
colnames(top_GO)[colnames(top_GO)=='Gene.Count'] <- "Gene Count"

#change ~ so the final table formats correctly:
top_GO <- data.frame(lapply(top_GO, function(x) {
  gsub("~"," ~ ",x)
}))
head(top_GO)

#clusters for genes higher in adult compared to P0 (the genes with positive LFC)
up_go <- top_GO %>% filter(Direction=="Up")

#clusters for genes higher in P0 compared to adult (the genes with negative LFC)
down_go <- top_GO %>% filter(Direction=="Down")
```

Table for Terms
```{r,results = "asis"}

#downregulated:
kbl(down_go[,4:9], caption = "GO Terms Upregulated in P0 Compared to Adult") %>% 
  pack_rows("Annotation Cluster 1, Enrichment Score = 30.78288", 1, 3) %>%
  pack_rows("Annotation Cluster 2, Enrichment Score = 25.77239", 4, 14) %>%
  pack_rows("Annotation Cluster 3, Enrichment Score = 25.33406", 15, 17) %>%
  kable_classic(full_width = F, html_font = "Cambria")

#upregulated:
kbl(up_go[,4:9], caption = "GO Terms Downregulated in P0 Compared to Adult") %>% 
  pack_rows("Annotation Cluster 1, Enrichment Score = 69.09126", 1, 7) %>%
  pack_rows("Annotation Cluster 2, Enrichment Score = 47.50504", 8, 24) %>%
  pack_rows("Annotation Cluster 3, Enrichment Score = 47.50504", 25, 27) %>%
  kable_classic(full_width = F, html_font = "Cambria")


```